---
title: "BIOSTAT 707 Checkpoint 1: cohort characterization and EDA"
author: "Yifei Chen"
date: now
format:
  html:
    embed-resources: true
    toc: true
engine: knitr
execute:
  echo: true
  warning: false
  error: false
---

```{r}
#| label: setup
#| message: false
library(data.table)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
here::i_am("checkpoint1.qmd")
data_dir <- here::here("data")
out_dir <- here::here("output")
dir.create(out_dir, showWarnings = FALSE)
expected_n <- 4000L

first_valid <- function(x) {
  x <- x[!is.na(x)]
  if (length(x)) x[[1]] else NA_real_
}
last_valid <- function(x) {
  x <- x[!is.na(x)]
  if (length(x)) x[[length(x)]] else NA_real_
}
min_valid <- function(x) if (all(is.na(x))) NA_real_ else min(x, na.rm = TRUE)
max_valid <- function(x) if (all(is.na(x))) NA_real_ else max(x, na.rm = TRUE)
mean_valid <- function(x) if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
```

# Data and analysis plan

This report describes the 4,000 ICU stays in Challenge 2012 Set A. Each record
contains observations from the first 48 hours after ICU admission. The outcome
file is joined for description, while its post-admission fields are kept separate
from candidate predictors. The input files come from the
[PhysioNet Challenge 2012 dataset](https://physionet.org/content/challenge-2012/1.0.0/)
(Silva et al., 2012; full citations below).
All file paths are relative to this repository. The analysis reads `data/` and
writes derived files only to `output/`.

```{r}
#| label: check-inputs
files <- list.files(
  file.path(data_dir, "set-a"), pattern = "^[0-9]+\\.txt$", full.names = TRUE
)
stopifnot(
  "Expected 4,000 Set A record files" = length(files) == expected_n,
  "Outcomes-a.txt is missing" = file.exists(file.path(data_dir, "Outcomes-a.txt"))
)
outcomes <- readr::read_csv(
  file.path(data_dir, "Outcomes-a.txt"), show_col_types = FALSE
)
stopifnot(
  "Expected 4,000 outcome rows" = nrow(outcomes) == expected_n,
  "Outcome IDs must be unique" = n_distinct(outcomes$RecordID) == expected_n,
  "Outcome labels must be binary" = all(outcomes$`In-hospital_death` %in% c(0, 1))
)
```

# The long table

Each record file has `Time`, `Parameter`, and `Value` columns. The filename and
the `RecordID` line within the file must agree. `RecordID` becomes a column, so
that identifying line is omitted from the measurement rows. The original value
is retained in `Value_raw`; `Value` is the analysis value. PhysioNet defines
`-1` as missing or unknown and says valid values are nonnegative. Therefore all
negative values, including other negative codes, become `NA` in `Value`.

```{r}
#| label: build-long
read_record <- function(path) {
  x <- data.table::fread(path, showProgress = FALSE)
  if (!identical(names(x), c("Time", "Parameter", "Value"))) {
    stop("Unexpected columns in ", basename(path))
  }
  id <- as.integer(tools::file_path_sans_ext(basename(path)))
  embedded_id <- x[Parameter == "RecordID", Value]
  if (length(embedded_id) != 1L || embedded_id != id) {
    stop("RecordID mismatch in ", basename(path))
  }
  x[, RecordID := id]
  x
}
long <- data.table::rbindlist(lapply(files, read_record), use.names = TRUE) |>
  tibble::as_tibble() |>
  filter(Parameter != "RecordID") |>
  mutate(
    hour = as.numeric(substr(Time, 1, 2)) +
      as.numeric(substr(Time, 4, 5)) / 60,
    Value_raw = as.numeric(Value),
    Value = if_else(Value_raw < 0, NA_real_, Value_raw)
  ) |>
  select(RecordID, Time, hour, Parameter, Value_raw, Value)
stopifnot(
  "Some times could not be parsed" = !anyNA(long$hour),
  "Times must be within the first 48 hours" = all(between(long$hour, 0, 48)),
  "Record IDs differ from outcomes" = setequal(unique(long$RecordID), outcomes$RecordID)
)
readr::write_csv(long, file.path(out_dir, "set-a_long.csv"))
```

The long table contains `r format(nrow(long), big.mark = ",")` measurement
rows from `r n_distinct(long$RecordID)` stays and
`r n_distinct(long$Parameter)` measurement parameters. There are
`r format(sum(long$Value_raw == -1), big.mark = ",")` explicit `-1` entries and
`r sum(long$Value_raw < 0 & long$Value_raw != -1)` other negative entries.

# Data quality and decisions

The following table shows the negative values other than `-1`. Because the
source documentation says valid measurements are nonnegative, these values are
also treated as missing in `Value`, while `Value_raw` preserves the evidence.

```{r}
#| label: negative-values
invalid_negative <- long |>
  filter(Value_raw < 0, Value_raw != -1) |>
  count(Parameter, Value_raw, name = "Rows", sort = TRUE)
knitr::kable(invalid_negative, caption = "Negative values other than the documented -1 code")
```

The next summaries expose unusual nonnegative values. The ranges are *review
flags*, not automatic deletion rules: severe illness and measurement error can
both produce extreme readings. We retain these values in the derived tables and
make that choice explicit. An individual value needs clinical and source-data
review before a later modeling analysis decides whether to exclude it.

```{r}
#| label: parameter-ranges
parameter_ranges <- long |>
  group_by(Parameter) |>
  summarise(
    rows = n(), missing_or_invalid = sum(is.na(Value)),
    minimum = min_valid(Value), median = if (all(is.na(Value))) NA_real_ else median(Value, na.rm = TRUE),
    maximum = max_valid(Value), .groups = "drop"
  ) |>
  arrange(Parameter)
knitr::kable(parameter_ranges, digits = 2,
             caption = "Observed ranges after documented and invalid negative values are set to missing")
```

```{r}
#| label: plausibility-flags
bounds <- tibble::tribble(
  ~Parameter, ~lower, ~upper,
  "Age", 18, 120,
  "Height", 100, 230,
  "Weight", 20, 400,
  "HR", 20, 250,
  "RespRate", 4, 60,
  "Temp", 30, 43,
  "pH", 6.8, 7.8,
  "GCS", 3, 15,
  "FiO2", 0, 1,
  "SaO2", 50, 100,
  "Gender", 0, 1,
  "ICUType", 1, 4
)
flags <- long |>
  inner_join(bounds, by = "Parameter") |>
  filter(!is.na(Value), Value < lower | Value > upper) |>
  group_by(Parameter) |>
  summarise(flagged_rows = n(), smallest = min(Value), largest = max(Value),
            .groups = "drop") |>
  arrange(desc(flagged_rows))
knitr::kable(flags, digits = 2,
             caption = "Values outside broad screening ranges, retained for review")
```

These flags include values that may be recording errors, for example a zero
body temperature or a pH far below a plausible physiological range. They are
not silently recoded because the documentation gives no correction rule for
such nonnegative entries. Median summaries below are less sensitive to isolated
extremes than means, but the wide file retains both summaries for inspection.

# Missingness over time in the long table

An absent row is not marked `NA` in a patient's file. To represent missingness,
we ask whether each admission has **at least one valid measurement** of each
time-series parameter in each two-hour interval. The denominator is all
`r expected_n` admissions, including those with no row for that parameter.
The final observation at hour 48 belongs to the last interval.

```{r}
#| label: temporal-missingness
#| fig-width: 11
#| fig-height: 9
time_series_parameters <- sort(setdiff(
  unique(long$Parameter), c("Age", "Gender", "Height", "ICUType")
))
temporal <- long |>
  filter(Parameter %in% time_series_parameters, !is.na(Value)) |>
  mutate(bin = pmin(floor(hour / 2), 23L)) |>
  distinct(RecordID, Parameter, bin) |>
  count(Parameter, bin, name = "admissions_measured")
temporal <- tidyr::expand_grid(
  Parameter = time_series_parameters, bin = 0:23
) |>
  left_join(temporal, by = c("Parameter", "bin")) |>
  mutate(
    admissions_measured = replace_na(admissions_measured, 0L),
    proportion_measured = admissions_measured / expected_n,
    Parameter = factor(Parameter, levels = rev(time_series_parameters))
  )
ggplot(temporal, aes(x = bin * 2, y = Parameter, fill = proportion_measured)) +
  geom_tile() +
  scale_x_continuous(breaks = seq(0, 46, by = 4), expand = c(0, 0)) +
  scale_fill_viridis_c(labels = scales::label_percent(), limits = c(0, 1)) +
  labs(title = "Measurement availability in the first 48 hours",
       subtitle = "Share of all 4,000 admissions with a valid reading in each two-hour interval",
       x = "Hours since ICU admission (interval start)", y = NULL,
       fill = "Measured") +
  theme_minimal(base_size = 10)
```

This graph describes the *measurement process*, not just explicit `-1` codes.
Variables with sparse or changing coverage may reflect clinical decisions about
testing as well as gaps in recording. The map does not establish why a
measurement is absent.

# The wide table

The wide table has one row per admission. Admission descriptors come from the
`00:00` entries. For each time-series parameter, it includes count, first,
last, minimum, maximum, and mean of valid readings in **0 to under 24 hours**
and **24 through 48 hours**. This preserves some change over time. A count of
zero means the variable was not validly measured in that window; its other
summaries remain missing. No imputation is performed.

```{r}
#| label: build-wide
static_parameters <- c("Age", "Gender", "Height", "ICUType", "Weight")
static <- long |>
  filter(hour == 0, Parameter %in% static_parameters) |>
  group_by(RecordID, Parameter) |>
  summarise(Value = first_valid(Value), .groups = "drop") |>
  pivot_wider(names_from = Parameter, values_from = Value)
static <- tibble::tibble(RecordID = outcomes$RecordID) |>
  left_join(static, by = "RecordID")

series_summaries <- long |>
  filter(Parameter %in% time_series_parameters) |>
  mutate(window = if_else(hour < 24, "h00_24", "h24_48")) |>
  arrange(RecordID, Parameter, hour) |>
  group_by(RecordID, Parameter, window) |>
  summarise(
    count = sum(!is.na(Value)),
    first = first_valid(Value), last = last_valid(Value),
    min = min_valid(Value), max = max_valid(Value), mean = mean_valid(Value),
    .groups = "drop"
  )
series_complete <- tidyr::expand_grid(
  RecordID = outcomes$RecordID,
  Parameter = time_series_parameters,
  window = c("h00_24", "h24_48")
) |>
  left_join(series_summaries, by = c("RecordID", "Parameter", "window")) |>
  mutate(count = replace_na(count, 0L))
wide_measurements <- series_complete |>
  pivot_wider(
    names_from = c(Parameter, window),
    values_from = c(count, first, last, min, max, mean),
    names_glue = "{Parameter}_{.value}_{window}"
  )
wide <- static |>
  left_join(wide_measurements, by = "RecordID") |>
  left_join(outcomes, by = "RecordID") |>
  mutate(
    `SAPS-I` = na_if(`SAPS-I`, -1),
    SOFA = na_if(SOFA, -1),
    Length_of_stay = na_if(Length_of_stay, -1)
  )
stopifnot(
  "Wide table should have 4,000 rows" = nrow(wide) == expected_n,
  "Wide IDs must be unique" = n_distinct(wide$RecordID) == expected_n,
  "HR counts disagree with long table" =
    sum(wide$HR_count_h00_24 + wide$HR_count_h24_48) ==
      sum(long$Parameter == "HR" & !is.na(long$Value))
)
readr::write_csv(wide, file.path(out_dir, "set-a_wide.csv"))
```

The wide table contains `r nrow(wide)` admissions and `r ncol(wide)` columns.
We retained the outcome fields for descriptive comparisons. They must be
excluded from future prediction features. In particular, in-hospital death,
length of stay, and survival time are known only after the prediction window;
the supplied SAPS-I and SOFA scores should not be assumed to be usable
predictors without confirming their calculation time and inputs.

# Table for cohort characteristics

Table 1 summarizes admission characteristics and supplied severity scores
overall and by in-hospital death. Continuous entries use median and
interquartile range; categorical entries use count and percent. The severity
scores are descriptive here, not proposed model inputs.

```{r}
#| label: table1
table1_data <- wide |>
  mutate(
    Death = factor(if_else(`In-hospital_death` == 1, "Died", "Survived"),
                   levels = c("Survived", "Died")),
    Sex = factor(Gender, levels = c(0, 1), labels = c("Female", "Male")),
    ICU = factor(ICUType, levels = 1:4,
                 labels = c("Coronary care", "Cardiac surgery recovery",
                            "Medical ICU", "Surgical ICU"))
  ) |>
  select(Death, Age, Sex, ICU, Height, Weight, `SAPS-I`, SOFA)
gtsummary::tbl_summary(
  table1_data,
  by = Death,
  statistic = list(
    gtsummary::all_continuous() ~ "{median} ({p25}, {p75})",
    gtsummary::all_categorical() ~ "{n} ({p}%)"
  ),
  missing = "ifany"
) |>
  gtsummary::add_overall(last = FALSE)
```

# Outcome summary

In-hospital death is the binary outcome: 1 means death during the hospital
stay. The `Survival` field uses `-1` when no death was recorded, so it is not
interpreted as an ordinary negative duration.

```{r}
#| label: outcomes
n_deaths <- sum(wide$`In-hospital_death` == 1)
death_rate <- n_deaths / nrow(wide)
los <- wide$Length_of_stay
outcome_table <- tibble::tibble(
  Measure = c("Admissions", "In-hospital deaths", "Survivors",
              "In-hospital mortality", "Length of stay, median (IQR), days",
              "Length of stay, range, days"),
  Result = c(
    format(nrow(wide), big.mark = ","), format(n_deaths, big.mark = ","),
    format(nrow(wide) - n_deaths, big.mark = ","),
    sprintf("%.1f%%", 100 * death_rate),
    sprintf("%.1f (%.1f, %.1f)", median(los, na.rm = TRUE),
            quantile(los, 0.25, na.rm = TRUE), quantile(los, 0.75, na.rm = TRUE)),
    sprintf("%.0f to %.0f", min(los, na.rm = TRUE), max(los, na.rm = TRUE))
  )
)
knitr::kable(outcome_table)
```

# Missingness in the wide table

A missing summary mean means no valid reading in that parameter and window.
The count columns deliberately hold zero instead of `NA`; plotting those
columns as missing would hide the measurement pattern. The first plot ranks
missing proportions for the mean summaries. The second displays selected
features across a reproducible sample of 200 admissions.

```{r}
#| label: wide-missingness-rates
#| fig-width: 9
#| fig-height: 8
mean_columns <- grep("_mean_h", names(wide), value = TRUE)
missing_rates <- tibble::tibble(
  feature = mean_columns,
  missing_fraction = colMeans(is.na(wide[mean_columns]))
) |>
  arrange(desc(missing_fraction))
ggplot(head(missing_rates, 30),
       aes(x = missing_fraction, y = reorder(feature, missing_fraction))) +
  geom_col(fill = "#3265a8") +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(title = "Thirty most frequently absent patient-window summaries",
       x = "Admissions without a valid measurement", y = NULL) +
  theme_minimal(base_size = 10)
```

```{r}
#| label: wide-missingness-map
#| fig-width: 11
#| fig-height: 7
selected_parameters <- c("HR", "Temp", "RespRate", "Glucose", "Creatinine",
                         "Lactate", "Albumin", "Platelets", "pH", "FiO2",
                         "TroponinI", "TroponinT")
selected_columns <- intersect(
  as.vector(outer(selected_parameters, c("h00_24", "h24_48"),
                  function(p, w) paste0(p, "_mean_", w))),
  names(wide)
)
set.seed(707)
sample_ids <- sample(wide$RecordID, min(200L, nrow(wide)))
wide_map <- wide |>
  filter(RecordID %in% sample_ids) |>
  arrange(RecordID) |>
  select(RecordID, all_of(selected_columns)) |>
  mutate(admission_index = row_number()) |>
  pivot_longer(all_of(selected_columns), 
                names_to = "feature", values_to = "value") |>
  mutate(
    feature = factor(feature, levels = rev(selected_columns)),
    status = if_else(is.na(value), "Missing", "Measured")
  )
ggplot(wide_map, aes(x = admission_index, y = feature, fill = status)) +
  geom_tile() +
  scale_fill_manual(values = c(Missing = "#d9dce3", Measured = "#216e8c")) +
  labs(title = "Missingness map of selected patient-window summaries",
       x = "Sampled admissions, ordered by RecordID", y = NULL, fill = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank())
```

The different patterns between the two windows matter: a variable may be
measured only early, only late, or not at all. We preserve those distinctions
and do not fill absent summaries with an overall mean at this exploratory
stage.

## Is the missingness informative?

The following descriptive comparison asks whether in-hospital mortality differs
between stays where a variable was ever validly measured and stays where it
was not. It does not establish that measuring the variable changes mortality:
the ordering decision may reflect patient condition and clinical practice.

```{r}
#| label: measurement-and-death
chosen_parameters <- c("HR", "Lactate", "Albumin", "TroponinI")
ever_measured <- long |>
  filter(Parameter %in% chosen_parameters, !is.na(Value)) |>
  distinct(RecordID, Parameter) |>
  mutate(measured = TRUE)
measurement_death <- tidyr::expand_grid(
  RecordID = outcomes$RecordID, Parameter = chosen_parameters
) |>
  left_join(ever_measured, by = c("RecordID", "Parameter")) |>
  mutate(measured = replace_na(measured, FALSE)) |>
  left_join(outcomes |> select(RecordID, `In-hospital_death`), by = "RecordID") |>
  group_by(Parameter, measured) |>
  summarise(
    admissions = n(), deaths = sum(`In-hospital_death` == 1),
    death_rate = deaths / admissions, .groups = "drop"
  ) |>
  mutate(death_rate = scales::percent(death_rate, accuracy = 0.1))
knitr::kable(measurement_death,
             caption = "Mortality by whether a variable was ever measured")
```

# Understanding the eventual Challenge assessment

No model is fitted or scored for Checkpoint 1. For a future binary mortality
prediction, sensitivity is $TP/(TP+FN)$ and positive predictive value is
$TP/(TP+FP)$. The Challenge's first criterion is the **smaller** of these two
numbers, so a high value of only one is insufficient. Its second criterion
assesses predicted risks: patients are grouped into risk deciles and a modified
Hosmer–Lemeshow discrepancy compares observed and expected deaths. The original
Challenge divides that discrepancy by the spread between mean risk in the top
and bottom deciles ($H/D$). These criteria concern future predictions, not
the descriptive work in this report. See the
[Challenge scoring description](https://physionet.org/content/challenge-2012/1.0.0/).

# AI-use statement

I used AI as an editing and learning aid while working with
R and Quarto. Because `knitr` was new to me, I asked how R chunks are run and
how their results appear in a rendered document. AI helped me understand less
familiar choices such as `here` for repository-relative paths and chunk labels
such as `#| label: check-inputs`. It also helped organize the *Data and analysis
plan* section. I used AI to refine some descriptions that initially sounded
too conversational and did not clearly connect the findings to the displayed
results. The revised wording is more formal and states the observed patterns
more precisely.

For the missingness map, I used AI feedback to review the tile layout,
figure size, labels, theme, and colors, including the `geom_tile()` and
`scale_fill_manual()` settings. AI also suggested retaining an additional
bar chart because it ranks the widest gaps in measurement coverage across
all variable-window summaries, whereas the tile map shows patterns for
individual admissions. After discussing that distinction, I accepted the
extra graph. When the report did not render, I used AI to
inspect the error message and identify the Quarto cache issue. These
interactions helped me understand unfamiliar tools and improve the report's
readability. A limitation is that suggestions can introduce complexity or
overlook details in the data. To review the AI-generated code, I asked for 
explanations of unfamiliar R and Quarto choices, including here, knitr chunk 
labels, and package startup messages. I ran the report with Pixi and used 
AI-assisted checks to compare variable names with the original data and 
inspect the resulting tables, graphs, and CSV files. Those checks 
confirmed 4,000 unique admissions and identified an incorrect troponin 
variable name, which was corrected. This review helped me catch implementation 
errors.

# Data citations

Silva, I., Moody, G., Mark, R. G., & Celi, L. A. (2012).
*Predicting Mortality of ICU Patients: The PhysioNet/Computing in Cardiology
Challenge 2012* (version 1.0.0) [Data set]. PhysioNet.
<https://physionet.org/content/challenge-2012/1.0.0/>.

Silva, I., Moody, G., Scott, D. J., Celi, L. A., & Mark, R. G. (2012).
Predicting in-hospital mortality of ICU patients: The PhysioNet/Computing in
Cardiology Challenge 2012. *Computing in Cardiology, 39*, 245–248.
<https://physionet.org/files/challenge-2012/1.0.0/papers/0245.pdf>.
